Observation of a Magnetic-Torque Enhanced by Interacting Spin Ensembles

### 2021-02-19 Mes papiers/Méca CR/18112020/Meca_cross.tex

Here, we employ resonant dipolar interactions between the spins of negatively charged nitrogen-vacancy centers to tune the orientation of a micro-mechanical oscillator. 
Specifically, we use NV centers inside a diamond that is levitating in a Paul trap and use resonant cross-relaxation between them to actuate
the spin-torque.
%We first present the results and conclude by discussing the potential applications behind our observations.
The key mechanism is depicted in Fig.~\ref{principle}. 
As can be seen in the left panel, NV centers can be found in four different orientations in the diamond crystalline structure. Due to quasi-rotational invariance of the problem, although each orientation could exert a significant torque to the diamond, the total spin-torque $\tau_s$ is very small.
%Our approach lies in the cross-relaxation (CR) that takes place when two spins with different orientations and polarization become resonant. 
%As depicted in Fig.~\ref{principle}-a), quasi-rotational invariance of the problem can make the sum of the spin torque of the four spin directions negligible. 
However by tuning an external magnetic field, resonant dipole-dipole interactions between the spin of NV centers of different orientations can be enhanced. 
When they become resonant, the polarization of the different orientations can then be exchanged in a process called cross-relaxation.
The right panel of Fig.~\ref{principle}, shows a cross-relaxation mechanism that partly removes the contribution from two classes of NV centers (labelled 1 and 3 in Fig.~\ref{principle}), which breaks the four-spin rotational invariance. The total spin torque $\tau_s$ can then be large enough to rotate the diamond.

%We first show that cross-relaxation can be effective in micro-diamonds that are heavily doped by NV centers.
%The NV center has two unpaired electrons in its ground state. In the triplet state, the dipolar interaction between them leads to a zero-field splitting $D=2.87$ GHz.
%The three spin states are labelled $| m_s={0}, {\pm 1}\rangle$.
%Due to an intersystem crossing in the optically excited state of the NV centers \cite{Doherty}, the state $m_s=\ket{0}$ is brighter than the state $m_s=\ket{\pm 1}$ upon green laser optical illumination \cite{Hopper}. 
%This provides a means to read out the Zeeman splitting by scanning a microwave tone around the resonance.
The diamonds that we use are in the form of a powder of particles that have a diameter of 15 $\mu$m. They are supplied by the company Adamas, which produces diamonds with a concentration of NV centers in the 3-4 ppm range. %The dipolar interactions may thus be effective since the average distance between the NV centers of about 10 nm. 
We load the micro-diamonds in a Paul trap that is similar to the one used in \cite{delordPRL} and depicted in Fig. \ref{setup}-a). One important feature of this system is the angular confinement provided by the trap, which offers the unique possibility to study the spin-induced torques.
Also, this geometry enables to apply a microwave signal directly on the trapping electrode, which provides an efficient means to excite the spins.
The photoluminescence from the NV centers is detected using standard confocal microscopy. We use about 100 $\mu$W of laser light at 532nm to polarize the NV centers. The laser is focused via a lens inside a vacuum chamber which has a numerical aperture of 0.5 and a working distance of 8mm. The focal point of the laser is kept few tens of micrometers away from the micro-diamond to mitigate the effect of radiation pressure and to enable laser excitation of the whole diamond \cite{delordPRL,delord2016}.

@@

Such a process has not been presented with nano- or micro-particles in the literature to the best of our knowledge. Smaller diamond particles in fact tend to suffer from extra parasitic surface effects such as spin depolarization due to interaction with paramagnetic dangling bonds on the surface \cite{Tetienne}, or enhanced charge transfer between the NV$^0$ and NV$^-$ charge states \cite{Dhomkar} so it is essential to verify that it can be observed with micro-particles. 
We start by searching for CR using micro-diamonds that are physically attached to the trap, by employing a fixed bias magnetic field $||\bf B_{\rm bias}||\approx$100 G and by tuning another magnetic field $\bf B_{\rm em}$ at some angle with $\bf B_{\rm bias}$ using an electromagnet (see Fig.~\ref{setup}-b)).

At specific magnetic field directions with respect to the crystalline axes, degeneracy between the spin of NV centers can be reached \cite{van_oort_cross-relaxation_1989}. %We check this systematically by searching for a reduction of the PL using APD$_2$ and using Opticall-Detected-Magnetic-Resonance (see SI) to correlated it with degeneracies. %We first characterize the depolarization of a diamond that is attached to trap ring to ensure that cross-relaxation can be observed in our samples.
%POUR LES SI 
%It is important to characterize this effect further when using smaller particles such micro-diamonds where surface effects can play a role. 
%where, to our knowledge, such cross-relaxations have not been observed. 
%Indeed, surface effects may start to play a role and change quantitatively.
In this case, the longitudinal decay time from NV centers should be reduced. To verify this, we measured the $T_1$ time by applying a green laser that polarizes the NV centers and measure the photoluminescence at a later time.
Such a measurement however turned out to be significantly impacted by recharging of NV centers in the dark \cite{choi_depolarization_2017, mrozek_longitudinal_2015, giri_selective_2019, giri_coupled_2018}.
In order to accurately measure the $T_1$ and remove the changing PL due to the recharging effects, we use the sequence presented in the SI, where a microwave pulse is applied or not prior to spin relaxation. The PL signals acquired in the two different measurements are then subtracted and shown for different degeneracy configurations in Figure \ref{setup}-c).
In the absence of degeneracy, we observe an exponentially decaying profile, from which we extract a $T_1=1.08$ ms, already shorter than the phonon limited lifetime in dilute bulk materials \cite{Tetienne}. This lifetime is even further reduced to 0.28 ms (resp. 0.15 ms) when two (resp. four) orientations are brought to resonance. This hints towards the role played by dipolar interactions, which are enhanced when classes of NV centers are resonant \cite{van_oort_cross-relaxation_1989, choi_depolarization_2017}. The effective density of NV being able to exchange spin quanta indeed increases, resulting in a stronger depolarization process.

@@

The main goal our study is to search for the mechanical action of such dipolar induced relaxations with diamonds levitating in the Paul trap.
%Following the picture presented in Fig. 1, we will seek for changes in the diamond orientation at CR conditions. 
%For a given levitating particle, we can optimize, in real time, the signal coming from the angular displacement of the particle by selecting the most favorable region of the particle image. To do this, we monitor our optical signal while switching a microwave field tuned to one of the ESR transitions, at a frequency of 1 Hz.  Alignment is realized by maximizing the variation of the coupled light as the diamond rotates between two angular positions. 
One major basic ingredient is the spin-torque coming from the NV centers when they are polarized in the ground state. 
Let us consider first the dependence of the ground state energy of a single spin as a function of the angle between a magnetic field and the NV axis.
The Hamiltonian for one NV orientation with quantization axis $z$ in the particle frame reads 
\begin{equation}\hat{H}_{\rm NV}=\hbar D \hat{S}_z^2+ \hbar \gamma_e \bf B  \cdot \bf\hat S,
\end{equation}
where $\bf\hat S$ is the spin-vector, $D=(2\pi)2.87$ GHz the zero-field splitting and $\bf B$ is the external magnetic field.
Under the condition $\gamma ||\bm B|| \ll D$, and assuming an NV center and a magnetic field in an $(x,z)$ plane, $H_{B}=  \hbar \gamma_e {\bf B}  \cdot {\bf \hat S}=\hbar\gamma_e B  ( S_x \sin\theta + S_z \cos\theta)$ can be treated as a perturbation to the anisotropic part $\hbar D \hat{S}_z^2$ of the Hamiltonian. Here, $\theta$ is the angle between the magnetic field and the body-fixed NV center axis $z$.
The perturbed energy $\epsilon_g$ of the ground state due to the transverse B field $B_\perp=B \sin(\theta)$ is then
\begin{equation} \epsilon_g=\sum_{m_s=\pm 1}\frac{ |\bra{0} H_B \ket{\pm 1}|^2}{-\epsilon^0_{\pm 1}}=-\hbar\frac{(\gamma_e B_\perp)^2 }{D}
\end{equation}
\begin{figure*}[!ht]
  \centering \scalebox{0.45}{\includegraphics{CRmeca}}
  \caption{a) Top :  Signal reflected off the diamond surface as a function of microwave frequency for three different magnetic field values. Bottom : sketch showing the crossing of a crystal plane when the magnetic field angle is tuned.
b) PL detection as a function of B$_{\rm em}$ across a dipole-dipole resonance. i) experimental data with gaussian fit, ii) simulation of the population in $\ket{m_s=0}$ state, taking into account cross-relaxation (plain) or not (dashed).
c) Angular detection as a function of B$_{em}$ across a co-resonance. i) experimental data with gaussian fit, ii) simulation of the magnetic torque applied to the diamond, taking into account cross-relaxation (plain) or not (dashed).
  }
  \label{data}
\end{figure*}
A direct use of the Feynman-Hellman theorem can give the torque in the ground state. We find that 
\begin{equation}\tau_s=- \frac{\partial \epsilon_g}{\partial \theta}=\hbar  \frac{(\gamma_e B)^2}{D} \sin 2\theta.\label{Eq1}
\end{equation}
A proof of the applicability of this theorem is presented in the SI. 
At an angle $\theta=\pi/4$, where the torque is maximized and at a B field of 100 G, we find $\tau_s\approx 5\times 10^{-17}$ N.m. Using $10^9$ spins polarized in the ground state and taking a librational confinement frequency of the diamond in the Paul trap to be around $\omega_\theta/(2\pi)=1$ kHz, we obtain an angular displacement 
of $\tau/I\omega_\theta^2\approx$1 mrad, which can be measured with a high signal-to-noise ratio in our set-up \cite{DelordNat}.
%We checked numerically that at these magnetic field values and angle, the NV are still polarized mostly in the ground state. 
%Note that the torque is zero at $\theta=mod (\pi/2)$ where the eigenstates of $S_z$ are not mixed. 
%The Hamiltonian ground state is the eigenstate $| S=1,m_s=0\rangle$, which is is non magnetic.

@@

As already hinted to in the explanations of Fig. 1, the contributions from the other NV classes must also be taken into account.
Fig. \ref{Numerics} presents the result of numerical calculations of the torque coming from the four classes of NV centers.   
Fig. \ref{Numerics}-a) shows the torque magnitude as a function of $\theta$ and $\phi$ without taking into account the cross-relaxations. The torque from each of the four classes appear clearly from the symmetry. Their different contributions however sum up to give a maximum torque of around $10^{-28}$ N.m, which is 20 times smaller than the torque that can be obtained for a single class due to the quasi-rotational invariance of the 4-NV classes. When two classes of NV center are resonant however, the induced cross-relaxation partly breaks this rotational invariance. 
Fig. \ref{Numerics}-b) shows the same plot, but including the effect of the cross-relaxation when classes of NV centers are degenerate (see SI).  
Here we use numbers that are deduced from our experimental observations of the CR induced change of the $T_1$.
One can see that a new pattern with larger spin-torque values is superimposed to the previous map. These larger values correspond to crossings of the crystal planes where NV degeneracies occur. 
At these coordinates, one recovers the torque estimation of Eq.~\ref{Eq1}, found for a single class, which would then imply a strong enough torque for overcoming the Paul trap confinement.

@@

To observe the effect of such resonant dipolar interactions experimentally, we use similar parameters and magnetic field arrangement than when the diamonds were not levitating. The diamond crystalline direction with respect to the magnetic field direction is characterized with greater precision than ODMR by recording the spin-torques given to the diamond when a microwave drives the spin to the $m_s=-1$ state. 
The angle detection is performed by collecting the back-reflected green light from the diamond interface (see Fig. 2), separated from the excitation light using a beam splitter. 
Fig. \ref{data}-a) shows spin-mechanical detection of spin-resonances for three different $\bf B_{\rm em}$ amplitudes. 
At 10 and 25 G, one can observe 4 peaks in the spectrum that demonstrate spin induced torque on the diamond from the 4 classes of NVs.
At 17 G however, two classes merge at 2.75 GHz. This is where we expect to observe CR. 
%Note that perfect degeneracy would still make the diamond rotate when performing microwave scans since the microwave polarization is not the same for the two classes, and thus does not drive equally the two spin transitions.

Analysis (SI) suggest that here, since we observe a single degeneracy, the magnetic field crosses a plane that is perpendicular to the $(110)$ direction as shown in Fig. \ref{data}-a).
Fig. \ref{data}-b) shows the photoluminescence as a function of $\bf B_{\rm em}$ both experimentally (trace i) and numerically (trace ii).
As expected, the PL decreases across the degeneracies at around the same magnetic field value, as a consequence of the CR-induced reduction of the $T_1$ time.
Fig. \ref{data}-c), trace i) is a measurement of the diamond angular position acquired simultaneously to the PL. Trace ii) is the corresponding calculation.   
A pronounced variation of the reflected signal is also observed, demonstrating here the close correspondence between degeneracy and diamond rotation and letting us conclude that we are observing a significant change in the spin-torque as the dipolar interactions between the spins increase. Note that as opposed to the PL detection, the signal coming from the particle surface can increase or decrease depending on how the speckle is aligned to the fiber. This explains the differing shapes of the signals in the experiments and the simulations. 
Fitting this curve by a Gaussian, we deduce a width that is similar to the PL width (1.3G and 1.7G respectively). This gives a width of X MHz ? comparable to ???
Similar experiments were realized on different particles under different degeneracies and similar effects were observed.
In the SI, we show similar results taken under different bias magnetic fields and under a two-fold degeneracy. 
%Note that the present effect cannot be attributed to a change in the angular momentum, as in the Einstein de Haas effect. 
%The latter would manifest itself for much smaller particles sizes. Further, the rotation would occur dynamically and not manifest itself as a static torque as is the case in the present experiment.

>>>>

Here, we employ resonant dipolar interactions between the spins of negatively charged nitrogen-vacancy centers to tune the orientation of a micro-mechanical oscillator. 
Specifically, we use NV centers inside a diamond that is levitating in a Paul trap and use resonant cross-relaxation between them to actuate
the spin-torque.
%We first present the results and conclude by discussing the potential applications behind our observations.
The key mechanism is depicted in Fig.~\ref{principle}. 
As can be seen in the left panel, NV centers can be found in four different orientations in the diamond crystalline structure. Due to quasi-rotational invariance of the problem, although each orientation could exert a significant torque to the diamond, the total spin-torque $\tau_s$ is very small.
%Our approach lies in the cross-relaxation (CR) that takes place when two spins with different orientations and polarization become resonant. 
%As depicted in Fig.~\ref{principle}-a), quasi-rotational invariance of the problem can make the sum of the spin torque of the four spin directions negligible. 
However by tuning an external magnetic field, resonant dipole-dipole interactions between the spin of NV centers of different orientations can be enhanced. 
When they become resonant, the polarization of the different orientations can then be exchanged in a process called cross-relaxation. The conditions on the magnetic field for a cross-relaxation to happen are described in Sec. I of the Supplementary Material (SM)\cite{SM_CR_meca}. \nocite{DelordPRL} \nocite{delordPhD} \nocite{DelordNat} \nocite{qutip1} \nocite{qutip2} \nocite{van_oort_cross-relaxation_1989} \nocite{choi_depolarization_2017}
The right panel of Fig.~\ref{principle}, shows a cross-relaxation mechanism that partly removes the contribution from two classes of NV centers (labelled 1 and 3 in Fig.~\ref{principle}), which breaks the four-spin rotational invariance. The total spin torque $\tau_s$ can then be large enough to rotate the diamond.

%We first show that cross-relaxation can be effective in micro-diamonds that are heavily doped by NV centers.
%The NV center has two unpaired electrons in its ground state. In the triplet state, the dipolar interaction between them leads to a zero-field splitting $D=2.87$ GHz.
%The three spin states are labelled $| m_s={0}, {\pm 1}\rangle$.
%Due to an intersystem crossing in the optically excited state of the NV centers \cite{Doherty}, the state $m_s=\ket{0}$ is brighter than the state $m_s=\ket{\pm 1}$ upon green laser optical illumination \cite{Hopper}. 
%This provides a means to read out the Zeeman splitting by scanning a microwave tone around the resonance.
The diamonds that we use are in the form of a powder of particles that have a diameter of 15 $\mu$m. They are supplied by the company Adamas, which produces diamonds with a concentration of NV centers in the 3-4 ppm range. %The dipolar interactions may thus be effective since the average distance between the NV centers of about 10 nm. 
We load the micro-diamonds in a Paul trap that is similar to the one used in \cite{DelordPRL} and depicted in Fig. \ref{setup}-a). One important feature of this system is the angular confinement provided by the trap, which offers the unique possibility to study the spin-induced torques.
Also, this geometry enables to apply a microwave signal directly on the trapping electrode, which provides an efficient means to excite the spins.
The photoluminescence from the NV centers is detected using standard confocal microscopy. We use about 100 $\mu$W of laser light at 532nm to polarize the NV centers. The laser is focused via a lens inside a vacuum chamber which has a numerical aperture of 0.5 and a working distance of 8mm. The focal point of the laser is kept few tens of micrometers away from the micro-diamond to mitigate the effect of radiation pressure and to enable laser excitation of the whole diamond \cite{DelordPRL,delord2016}.

@@

Such a process has not been presented with nano- or micro-particles in the literature to the best of our knowledge. Smaller diamond particles in fact tend to suffer from extra parasitic surface effects such as spin depolarization due to interaction with paramagnetic dangling bonds on the surface \cite{Tetienne}, or enhanced charge transfer between the NV$^0$ and NV$^-$ charge states \cite{Dhomkar} so it is essential to verify that it can be observed with micro-particles. 
We start by searching for CR using micro-diamonds that are physically attached to the trap, by employing a fixed bias magnetic field $||\bf B_{\rm bias}||\approx$100 G and by tuning another magnetic field $\bf B_{\rm em}$ at some angle with $\bf B_{\rm bias}$ using an electromagnet (see Fig.~\ref{setup}-a)). The change in orientation of the total magnetic field can be visualized in Sec. I of the SM \cite{SM_CR_meca}.

At specific magnetic field directions with respect to the crystalline axes, degeneracy between the spin of NV centers can be reached \cite{van_oort_cross-relaxation_1989}. %We check this systematically by searching for a reduction of the PL using APD$_2$ and using Opticall-Detected-Magnetic-Resonance (see SI) to correlated it with degeneracies. %We first characterize the depolarization of a diamond that is attached to trap ring to ensure that cross-relaxation can be observed in our samples.
%POUR LES SI 
%It is important to characterize this effect further when using smaller particles such micro-diamonds where surface effects can play a role. 
%where, to our knowledge, such cross-relaxations have not been observed. 
%Indeed, surface effects may start to play a role and change quantitatively.
In this case, the longitudinal decay time from NV centers should be reduced. To verify this, we measured the $T_1$ time by applying a green laser that polarizes the NV centers and measure the photoluminescence at a later time.
Such a measurement however turned out to be significantly impacted by recharging of NV centers in the dark \cite{choi_depolarization_2017, mrozek_longitudinal_2015, giri_selective_2019, giri_coupled_2018}.
In order to accurately measure the $T_1$ and remove the changing PL due to the recharging effects, we use the sequence presented in Sec. II of the SM \cite{SM_CR_meca}, where a microwave pulse is applied or not prior to spin relaxation. The PL signals acquired in the two different measurements are then subtracted and shown for different degeneracy configurations in Figure \ref{setup}-c).
In the absence of degeneracy, we observe an exponentially decaying profile, from which we extract a $T_1=1.08$ ms, already shorter than the phonon limited lifetime in dilute bulk materials \cite{Tetienne}. This lifetime is even further reduced to 0.28 ms (resp. 0.15 ms) when two (resp. four) orientations are brought to resonance. This hints towards the role played by dipolar interactions, which are enhanced when classes of NV centers are resonant \cite{van_oort_cross-relaxation_1989, choi_depolarization_2017}. The effective density of NV being able to exchange spin quanta indeed increases, resulting in a stronger depolarization process.

@@

The main goal our study is to search for the mechanical action of such dipolar induced relaxations with diamonds levitating in the Paul trap.
%Following the picture presented in Fig. 1, we will seek for changes in the diamond orientation at CR conditions. 
%For a given levitating particle, we can optimize, in real time, the signal coming from the angular displacement of the particle by selecting the most favorable region of the particle image. To do this, we monitor our optical signal while switching a microwave field tuned to one of the ESR transitions, at a frequency of 1 Hz.  Alignment is realized by maximizing the variation of the coupled light as the diamond rotates between two angular positions. 
One major basic ingredient is the spin-torque coming from the NV centers when they are polarized in the ground state. 
Let us consider first the dependence of the ground state energy of a single spin as a function of the angle between a magnetic field and the NV axis.
The Hamiltonian for one NV orientation with quantization axis $z$ in the particle frame reads 
\begin{equation}\hat{H}_{\rm NV}=\hbar D \hat{S}_z^2+ \hbar \gamma_e \bf B  \cdot \bf\hat S,
\end{equation}
where $\bf\hat S$ is the spin-vector, $D=(2\pi)2.87$ GHz the zero-field splitting and $\bf B$ is the external magnetic field.
Under the condition $\gamma ||\bm B|| \ll D$, and assuming an NV center and a magnetic field in an $(x,z)$ plane, $H_{B}=  \hbar \gamma_e {\bf B}  \cdot {\bf \hat S}=\hbar\gamma_e B  ( S_x \sin\theta + S_z \cos\theta)$ can be treated as a perturbation to the anisotropic part $\hbar D \hat{S}_z^2$ of the Hamiltonian. Here, $\theta$ is the angle between the magnetic field and the body-fixed NV center axis $z$.
The perturbed energy $\epsilon_g$ of the ground state due to the transverse B field $B_\perp=B \sin(\theta)$ is then
\begin{equation} \epsilon_g=\sum_{m_s=\pm 1}\frac{ |\bra{0} H_B \ket{\pm 1}|^2}{-\epsilon^0_{\pm 1}}=-\hbar\frac{(\gamma_e B_\perp)^2 }{D}
\end{equation}
\begin{figure*}[!ht]
  \centering \scalebox{0.45}{\includegraphics{CRmeca}}
  \caption{a) Top :  Signal reflected off the diamond surface as a function of microwave frequency for three different magnetic field values. Bottom : sketch showing the crossing of a crystal plane when the magnetic field angle is tuned.
b) PL detection as a function of B$_{\rm em}$ across a dipole-dipole resonance. i) experimental data with gaussian fit, ii) simulation of the population in $\ket{m_s=0}$ state, taking into account cross-relaxation (plain) or not (dashed).
c) Angular detection as a function of B$_{em}$ across a co-resonance. i) experimental data with gaussian fit, ii) simulation of the magnetic torque applied to the diamond, taking into account cross-relaxation (plain) or not (dashed).
  }
  \label{data}
\end{figure*}
A direct use of the Feynman-Hellman theorem can give the torque in the ground state. We find that 
\begin{equation}\tau_s=- \frac{\partial \epsilon_g}{\partial \theta}=\hbar  \frac{(\gamma_e B)^2}{D} \sin 2\theta.\label{Eq1}
\end{equation}
A proof of the applicability of this theorem is presented in  Sec. III of the SM \cite{SM_CR_meca}. 
At an angle $\theta=\pi/4$, where the torque is maximized and at a B field of 100 G, we find $\tau_s\approx 5\times 10^{-17}$ N.m. Using $10^9$ spins polarized in the ground state and taking a librational confinement frequency of the diamond in the Paul trap to be around $\omega_\theta/(2\pi)=1$ kHz, we obtain an angular displacement 
of $\tau/I\omega_\theta^2\approx$1 mrad, which can be measured with a high signal-to-noise ratio in our set-up \cite{DelordNat}.
%We checked numerically that at these magnetic field values and angle, the NV are still polarized mostly in the ground state. 
%Note that the torque is zero at $\theta=mod (\pi/2)$ where the eigenstates of $S_z$ are not mixed. 
%The Hamiltonian ground state is the eigenstate $| S=1,m_s=0\rangle$, which is is non magnetic.

@@

As already hinted to in the explanations of Fig. 1, the contributions from the other NV classes must also be taken into account.
Fig. \ref{Numerics} presents the result of numerical calculations of the torque coming from the four classes of NV centers.   
Fig. \ref{Numerics}-a) shows the torque magnitude as a function of $\theta$ and $\phi$ without taking into account the cross-relaxations. The torque from each of the four classes appear clearly from the symmetry. Their different contributions however sum up to give a maximum torque of around $10^{-28}$ N.m, which is 20 times smaller than the torque that can be obtained for a single class due to the quasi-rotational invariance of the 4-NV classes. When two classes of NV center are resonant however, the induced cross-relaxation partly breaks this rotational invariance. 
Fig. \ref{Numerics}-b) shows the same plot, but including the effect of the cross-relaxation when classes of NV centers are degenerate. Details on the modelization can be found in Sec.V of the SM \cite{SM_CR_meca}.
Here we use numbers that are deduced from our experimental observations of the CR induced change of the $T_1$.
One can see that a new pattern with larger spin-torque values is superimposed to the previous map. These larger values correspond to crossings of the crystal planes where NV degeneracies occur. 
At these coordinates, one recovers the torque estimation of Eq.~\ref{Eq1}, found for a single class, which would then imply a strong enough torque for overcoming the Paul trap confinement.

@@

To observe the effect of such resonant dipolar interactions experimentally, we use similar parameters and magnetic field arrangement than when the diamonds were not levitating. The diamond crystalline direction with respect to the magnetic field direction is characterized with greater precision than ODMR by recording the spin-torques given to the diamond when a microwave drives the spin to the $m_s=-1$ state. 
The angle detection is performed by collecting the back-reflected green light from the diamond interface (see Fig. 2-a), separated from the excitation light using a beam splitter. 
Fig. \ref{data}-a) shows spin-mechanical detection of spin-resonances for three different $\bf B_{\rm em}$ amplitudes. 
At 10 and 25 G, one can observe 4 peaks in the spectrum that demonstrate spin induced torque on the diamond from the 4 classes of NVs.
At 17 G however, two classes merge at 2.75 GHz. This is where we expect to observe CR. 
%Note that perfect degeneracy would still make the diamond rotate when performing microwave scans since the microwave polarization is not the same for the two classes, and thus does not drive equally the two spin transitions.

Analysis developed in Sec.I of the SM \cite{SM_CR_meca} suggest that here, since we observe a single degeneracy, the magnetic field crosses a plane that is perpendicular to the $(110)$ direction as shown in Fig. \ref{data}-a).
Fig. \ref{data}-b) shows the photoluminescence as a function of $\bf B_{\rm em}$ both experimentally (trace i) and numerically (trace ii).
As expected, the PL decreases across the degeneracies at around the same magnetic field value, as a consequence of the CR-induced reduction of the $T_1$ time.
Fig. \ref{data}-c), trace i) is a measurement of the diamond angular position acquired simultaneously to the PL. Trace ii) is the corresponding calculation.   
A pronounced variation of the reflected signal is also observed, demonstrating here the close correspondence between degeneracy and diamond rotation and letting us conclude that we are observing a significant change in the spin-torque as the dipolar interactions between the spins increase. Note that as opposed to the PL detection, the signal coming from the particle surface can increase or decrease depending on how the speckle is aligned to the fiber. This explains the differing shapes of the signals in the experiments and the simulations. 
Fitting this curve by a Gaussian, we deduce a width that is similar to the PL width (2.1G and 2.8G respectively). This gives a width of 9 (resp. 12) MHz  comparable to the inhomogeneous broadening of the sample.
Similar experiments were realized on different particles under different degeneracies and similar effects were observed.
In Sec. IV of the SM, we show similar results taken under different bias magnetic fields and under a two-fold degeneracy. 
%Note that the present effect cannot be attributed to a change in the angular momentum, as in the Einstein de Haas effect. 
%The latter would manifest itself for much smaller particles sizes. Further, the rotation would occur dynamically and not manifest itself as a static torque as is the case in the present experiment.

### 2021-02-19 Mes papiers/Méca CR/18112020/SI/Meca_cross_SI.tex

\section{Experimental details}

>>>>

\section{Depolarization induced by NV-NV cross-relaxation}
As explained in the main text, when the density of NV$^-$ centers in the sample is high enough (typically for concentrations higher than 1 ppm), the ensemble of NV spins will lose some of its polarization through dipolar coupling between the NV centers. This phenomenon is at the heart of the mechanism that allows us to exalt the magnetic suceptibility of our diamond through dipolar interaction, and it has already been observed independently by many groups in bulk diamond %refs.

In particular, \cite{choi_depolarization_2017} proposes a model based on "fluctuators" : a subgroup of NV centers with a very short lifetime (possibly due to their electron tunnelling in and out of the NV site) that can act as a source of classical noise with a central frequency given by the usual transition frequencies of the NV center Hamiltonian. A prediction of this model is that the modified lifetime of the ensemble of NV centers should have a stretch exponential profile (of the form $\exp(-\sqrt{\frac{t}{T_1}})$) which we do observe experimentally.

\subsection{Stretch exponential profile of the lifetimes}
\begin{figure}
\label{stretch}
\includegraphics[scale=0.7]{T1-stretch_exp}
\caption{Lifetime measurement for a case where all four classes are degenerate. Plain orange line correspond to a stretch-exponential fit and dashed red line to a simple exponential fit}
\end{figure}
In the theory developed in \cite{choi_depolarization_2017}, the stretch exponential profile arises from the inhomogeneity of the distance from each NV centers to the closest fluctuators. If $\rho_{00}^s(t)$, the population in the $\ket{m_s=0}$ state for each NV centers, follow a law of the form $\rho_{00}^s(t) \propto \exp(-\gamma t)$, then, assuming an even spatial distribution of fluctuators, the authors of \cite{choi_depolarization_2017} show that the distribution in $\gamma$ should follow a law of the form $$\rho(\gamma) \approx \frac{e^{-\frac{1}{4 \gamma T_1}}}{\sqrt{4 \pi \gamma^3 T_1}} $$
Averaging then over all NV centers gives the stretch exponential profile observed from the ensemble :
$$ \rho_{00}^e(t) \propto \int_0^{+\infty} \rho(\gamma) e^{-\gamma t} d\gamma = e^{-\sqrt{\frac{t}{T_1}}} $$
where $ \rho_{00}^e(t)$ correspond to the average population in the $\ket{m_s=0}$ state for the ensemble of spins.

Fig. \ref{stretch} shows a lifetime measurement on a static micro diamond following the protocol described in sec.III. Here all four classes are resonant with the applied microwave frequency, which correspond to the maximum degree of degeneracy between the NV centers, and therefore the stronger modification of the lifetime induced by the resonant dipolar coupling. The signal we obtain was fitted using a stretch exponential profile  and a simple exponential profile. We can see that the stretch exponential profile ($R^2=0.981$) is in better agreement with the data than the exponential fit ($R^2=0.942$), in particular for the very short times (we expect the longer times to be dominated by the phonon-limited exponential lifetime).

Finally it should be noted that the stretch exponential profile arising from point-like depolarization sources is a relatively general result that has for example also been observed for the depolarization of NV centers induced by substitutional nitrogen (P1) defects in diamond \cite{Hall}

\subsection{•}

The easiest way to probe the mechanism of dipolar-induced modification of the lifetime, as well as to justify the fact that the effects come in fact from the NV-NV dipolar interaction, is to change the degeneracy conditions between the four classes of NV centers by tuning the magnetic field. (ref sec.I)
Because the NV spins can only exchange spin quanta when they are quasi-resonant, then by tuning the number of at degeneracy, you can modify the effective density of NV centers able to by reso
\section{Experimental details}
